---
title: "AAVSO Solar Composite"
subtitle: "Tidyverse port of COMPSOL_DB.PL v1.0.2 — Matthew Templeton, AAVSO HQ (2012-06-15)"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    self-contained: true
  pdf:
    toc: true
execute:
  echo: false
  warning: false
  message: false
params:
  month: 5        # Integer 1–12
  year:  2026     # Four-digit year ≥ 1940
  output_dir: "."
---

```{r setup, include=FALSE}
library(DBI)
library(RMariaDB)
library(tidyverse)
library(knitr)
library(kableExtra)
library(glue)
```

```{r constants}
K_DEFAULT <- 1.0    # fallback k-value when observer has none
W_DEFAULT <- 0.05   # fallback Scale/weight (compsol uses 0.05, not 0.1)

months_lut  <- c("XXX","Jan","Feb","Mar","Apr","May","Jun",
                 "Jul","Aug","Sep","Oct","Nov","Dec")

stopifnot(params$month >= 1L, params$month <= 12L, params$year >= 1940L)

monthyr_str <- paste0(months_lut[params$month], " ", params$year)
```

```{r connect}
dbh <- dbConnect(
  RMariaDB::MariaDB(),
  dbname   = "aavso",
  host     = "localhost",
  user     = "root",
  password = "dJj135790",
  port     = 3333
)
```

## Helper functions

```{r jd-helpers}
# Days before the start of each month in a non-leap year (month 1..12).
.nmon <- c(0L, 31L, 59L, 90L, 120L, 151L, 181L, 212L, 243L, 273L, 304L, 334L)

# Port of Perl julian_zero($month, $year)
julian_zero <- function(month, year) {
  nyr   <- as.integer(year)
  if (nyr < 100L) nyr <- nyr + 2000L
  nyr   <- nyr - 1601L
  dtime <- 2305813L
  while (nyr >= 400L) { nyr <- nyr - 400L; dtime <- dtime + 146097L }
  while (nyr >= 100L) { nyr <- nyr - 100L; dtime <- dtime + 36524L  }
  while (nyr >= 4L)   { nyr <- nyr - 4L;   dtime <- dtime + 1461L   }
  dtime <- dtime + 365L * nyr + .nmon[month]
  if (nyr == 3L && month > 2L) dtime <- dtime + 1L
  dtime
}

# Port of Perl old_solarjd($day, $month, $year, $ut)
# ut is a 4-character string "HHMM"; numeric values are zero-padded automatically.
# Returns Modified Julian Date (MJD = JD - 2,400,000.5).
old_solarjd <- function(day, month, year, ut) {
  ut_str <- str_pad(as.character(ut), 4L, side = "left", pad = "0")
  nhour  <- as.integer(substr(ut_str, 1L, 2L))
  nmin   <- as.integer(substr(ut_str, 3L, 4L))
  frac   <- ((60L * nhour) + nmin) / 1440.0
  frac + day + julian_zero(month, year) - 2400000.5
}

# Round-half-up — matches Perl int(x + 0.5)
round_half_up <- function(x, digits = 0L) {
  floor(x * 10^digits + 0.5) / 10^digits
}
```

## Fetch observations

```{r fetch-obs}
sun_data <- tbl(dbh, "sun_data") |>
  filter(
    monthyr == monthyr_str,
    revised == 0L,
    !Obs %in% c("BSJ1", "TMT", "WEO", "SSTW", "ARUB")
  ) |>
  collect() |>
  arrange(Obs, day, UT)

n_data      <- nrow(sun_data)
n_observers <- n_distinct(sun_data$Obs)

cat(sprintf("There were %d observations by %d observers.\n", n_data, n_observers))

if (n_data == 0L) {
  cat("No data found for", monthyr_str, "— nothing to do.\n")
  dbDisconnect(dbh)
  knitr::knit_exit()
}
```

## Fetch observer constants

compsol uses `revised = 0` when querying observer constants.

```{r fetch-obsconst}
olist <- unique(sun_data$Obs)

obsconst <- tbl(dbh, "sun_obsconst") |>
  filter(Obs %in% olist, revised == 0L) |>
  collect() |>
  select(Obs, k, Scale, name, updated) |>
  arrange(Obs, desc(updated)) |>
  slice_max(updated, by = Obs, n = 1L, with_ties = FALSE) |>
  mutate(
    k     = if_else(is.na(k)     | k     == 0, K_DEFAULT, k),
    Scale = if_else(is.na(Scale) | Scale == 0, W_DEFAULT, Scale),
    name  = coalesce(name, "OBS NOT IN OBSCONST!")
  )

kbl(
  obsconst |> select(Obs, name, k, Scale),
  caption  = glue("Observer constants used for {monthyr_str}"),
  digits   = 4,
  booktabs = TRUE
) |>
  kable_styling(full_width = FALSE)
```

## Compute per-observation quantities

```{r compute-obs}
obs <- sun_data |>
  left_join(obsconst |> select(Obs, k, Scale, name), by = "Obs") |>
  mutate(
    k      = coalesce(k,     K_DEFAULT),
    Scale  = coalesce(Scale, W_DEFAULT),
    name   = coalesce(name,  "OBS NOT IN OBSCONST!")
  ) |>
  rowwise() |>
  mutate(JD = old_solarjd(day, params$month, params$year, UT)) |>
  ungroup() |>
  mutate(
    kcount = round_half_up(W * k, 4L),
    # Integer day bucket — matches Perl: int($jd + 0.50001)
    lday   = floor(JD + 0.50001)
  )

# Warn about any observers not found in obsconst
missing_obs <- obs |>
  filter(name == "OBS NOT IN OBSCONST!") |>
  distinct(Obs)

if (nrow(missing_obs) > 0L) {
  cat(sprintf(
    "WARNING: k and Scale not found for observer(s): %s — defaults applied.\n",
    paste(missing_obs$Obs, collapse = ", ")
  ))
}
```

## Compute daily statistics

Daily raw mean (RAW) and weighted k-corrected mean (Ra) match the Perl
`compdaily` subroutine output columns.

```{r daily-stats}
daily_stats <- obs |>
  group_by(lday) |>
  summarize(
    numdat   = n(),
    # Arithmetic mean of raw Wolf number (dave[1] in Perl)
    raw_ave  = mean(W),
    # Weighted arithmetic mean of k-corrected Wolf number (dave[2] in Perl)
    kw_ave   = sum(Scale * kcount) / sum(Scale),
    # Unweighted sample std dev of W (dsig[1] in Perl)
    raw_sd   = if_else(n() > 1L, sd(W), 0),
    # Weighted sample std dev of kcount (dsig[2] in Perl)
    kw_var   = (sum(Scale * kcount^2) / sum(Scale)) - (sum(Scale * kcount) / sum(Scale))^2,
    kw_dfact = 1 - sum(Scale^2) / sum(Scale)^2,
    kw_sd    = if_else(n() > 1L, sqrt(replace_na(kw_var / kw_dfact, 0)), 0),
    .groups  = "drop"
  ) |>
  mutate(
    day_seq = row_number(),
    # Round-half-up to integer — matches Perl int($dave[1]+0.5)
    raw_rnd = as.integer(floor(raw_ave + 0.5)),
    kw_rnd  = as.integer(floor(kw_ave  + 0.5))
  ) |>
  select(day_seq, numdat, raw_ave, raw_sd, raw_rnd, kw_ave, kw_sd, kw_rnd, lday)
```

### Daily summary table

```{r daily-preview}
#| tbl-cap: "Daily statistics — RAW = arithmetic mean of W; Ra = weighted mean of k-corrected W"

daily_stats |>
  select(
    Day      = day_seq,
    `N obs`  = numdat,
    RAW      = raw_rnd,
    `RAW sd` = raw_sd,
    Ra       = kw_rnd,
    `Ra sd`  = kw_sd
  ) |>
  kbl(booktabs = TRUE, digits = 2, linesep = "") |>
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width    = FALSE
  )
```

## Observer tally

```{r obs-tally}
obs_tally <- obs |>
  count(Obs, name = "n_obs") |>
  left_join(obsconst |> select(Obs, name), by = "Obs") |>
  mutate(name = coalesce(name, "OBS NOT IN OBSCONST!")) |>
  arrange(Obs)

kbl(
  obs_tally,
  col.names = c("Obs", "Observations", "Name"),
  caption   = "Observation count by observer",
  booktabs  = TRUE
) |>
  kable_styling(full_width = FALSE)
```

## Write `soldata.txt` {#sec-write}

Output format matches the original Perl `printf` exactly.

```{r write-txt}
output_file <- file.path(params$output_dir, "soldata.txt")

# Averages over all days
n_days     <- nrow(daily_stats)
ave_numobs <- mean(daily_stats$numdat)
ave_raw    <- mean(daily_stats$raw_ave)
ave_kw     <- mean(daily_stats$kw_ave)

# --- body lines ---
body_lines <- daily_stats |>
  mutate(
    line = sprintf("%2i   %4i   %4i   %4i", day_seq, numdat, raw_rnd, kw_rnd)
  ) |>
  pull(line)

# --- footer lines ---
average_line <- sprintf(
  "Average  %6.1f   %6.1f   %6.1f",
  ave_numobs, ave_raw, ave_kw
)

tally_lines <- c(
  "Number of observations by observer:",
  "===================================",
  obs_tally |>
    mutate(line = sprintf("%-4s  %4i   %s", Obs, n_obs, name)) |>
    pull(line),
  "===================================",
  sprintf("No. of Observers: %3i",           n_distinct(obs$Obs)),
  "===================================",
  sprintf("Total No. of Observations: %4i",  n_data)
)

all_lines <- c(
  "DAY    NumObs    RAW     Ra",
  body_lines,
  average_line,
  tally_lines
)

writeLines(all_lines, output_file)
cat(sprintf(
  "Wrote %d day(s) + footer to: %s\n",
  n_days, output_file
))
```

```{r disconnect}
dbDisconnect(dbh)
cat("Database connection closed.\n")
```
