---
title: "AAVSO Solar Raw Processing"
subtitle: "Port of RAWPROC_DB.PL v1.0 — Matthew Templeton, AAVSO HQ (2012-06-13)"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    self-contained: true
  pdf:
    toc: true
execute:
  echo: true
  warning: false
  message: false
params:
  month: 3        # Integer 1–12
  year:  2026     # Four-digit year ≥ 1940
  output_dir: "."  # Directory for soldata.srt; set to your WD if needed
---

```{r}
#| label: setup
#| include: false

library(DBI)
library(RMySQL)      # swap for RMariaDB if using a MariaDB server
library(tidyverse)
library(stringr)
library(knitr)
library(kableExtra)
```

## Parameters

```{r}
#| label: params-validate

# Pull params and validate ---------------------------------------------------
month <- as.integer(params$month)
year  <- as.integer(params$year)

stopifnot(
  "month must be 1–12"    = month >= 1L & month <= 12L,
  "year must be ≥ 1940"   = year  >= 1940L
)

month_names <- c("Jan","Feb","Mar","Apr","May","Jun",
                 "Jul","Aug","Sep","Oct","Nov","Dec")
monthyr <- paste(month_names[month], year)   # e.g. "Mar 2026"

WD <- params$output_dir

cat("Processing:", monthyr, "\n")
```

Processing month-year: **`r monthyr`**

---

## Default fill-in values

Used when an observer has no `k` or weight (`Scale`) in `sun_obsconst`.
Keep consistent with historical values to avoid shifts in long-term statistics.

```{r}
#| label: defaults

K_DEFAULT <- 1.0
W_DEFAULT <- 0.1   # "Scale" / weight
```

---

## Database connection

Credentials are read from environment variables so they are never stored in
the document source.  Set them in your `.Renviron` or shell before rendering:

```
SOLAR_DB=solar
SOLAR_HOST=localhost
SOLAR_USER=youruser
SOLAR_PASS=yourpass
```

```{r}
#| label: db-connect

DB_NAME <- Sys.getenv("SOLAR_DB",   "solar")
DB_HOST <- Sys.getenv("SOLAR_HOST", "localhost")
DB_USER <- Sys.getenv("SOLAR_USER", "")
DB_PASS <- Sys.getenv("SOLAR_PASS", "")

dbh <- dbConnect(
  RMySQL::MySQL(),
  dbname   = DB_NAME,
  host     = DB_HOST,
  user     = DB_USER,
  password = DB_PASS
)
```

---

## Fetch observations {#sec-fetch-obs}

Pulls unrevised records for the selected month, excluding internal/test
observers `BSJ1`, `TMT`, and `WEO`.

```{r}
#| label: fetch-obs

sql_obs <- "
  SELECT uid, date, monthyr, day, see, UT,
         g, s, W, ng, sg, ns, ss, cg, cs,
         Obs, Remarks, header_id, submitted, revised
  FROM   sun_data
  WHERE  monthyr  = ?
    AND  revised  = 0
    AND  Obs NOT IN ('BSJ1','TMT','WEO')
  ORDER  BY Obs, day, UT
"
obs <- dbGetQuery(dbh, sql_obs, params = list(monthyr))
colnames(obs) <- c("uid","date","monthyr","day","see","UT",
                   "g","s","W","ng","sg","ns","ss","cg","cs",
                   "Obs","Remarks","header_id","submitted","revised")

n_obs       <- nrow(obs)
n_observers <- n_distinct(obs$Obs)
cat(sprintf("Retrieved %d observations from %d observer(s).\n",
            n_obs, n_observers))

# Early exit if no data — stops rendering cleanly without an error page
if (n_obs == 0L) {
  cat("No data found for", monthyr, "— nothing to do.\n")
  dbDisconnect(dbh)
  knitr::knit_exit()
}
```

**`r n_obs`** observations from **`r n_observers`** observer(s) retrieved for `r monthyr`.

---

## Fetch observer constants {#sec-obs-const}

```{r}
#| label: fetch-obsconst

olist <- unique(obs$Obs)

sql_const <- sprintf(
  "SELECT k, Scale, name, Obs, AAVSO_obscode, SIDC_submitter, updated
   FROM   sun_obsconst
   WHERE  Obs IN (%s)",
  paste(rep("?", length(olist)), collapse = ",")
)
obsconst <- dbGetQuery(dbh, sql_const, params = as.list(olist))
colnames(obsconst) <- c("k","Scale","name","Obs","AAVSO","SIDC","updated")

# Apply defaults for missing / zero values
obsconst <- obsconst |>
  mutate(
    k     = if_else(is.na(k)     | k     == 0, K_DEFAULT, k),
    Scale = if_else(is.na(Scale) | Scale == 0, W_DEFAULT, Scale)
  )
```

```{r}
#| label: tbl-obsconst
#| tbl-cap: "Observer constants used in this run"

obsconst |>
  select(Obs, name, k, Scale, AAVSO, SIDC, updated) |>
  arrange(Obs) |>
  kbl(booktabs = TRUE, linesep = "") |>
  kable_styling(latex_options = c("striped","hold_position"),
                full_width    = FALSE)
```

---

## Julian Date helpers {#sec-jd}

Direct port of the original JULIAN0 QuickBasic algorithm used in the Perl
script. Results are Modified Julian Dates (MJD).

```{r}
#| label: jd-functions

# julian_zero(): MJD base for day 0 of the given month/year
julian_zero <- function(nmo, nyear) {
  nmon <- c(0L, 0L, 31L, 59L, 90L, 120L, 151L, 181L, 212L, 243L, 273L, 304L, 334L)
  nyr  <- if (nyear < 100L) nyear + 2000L else nyear
  nyr  <- nyr - 1601L

  dtime <- 2305813L
  while (nyr >= 400L) { nyr <- nyr - 400L; dtime <- dtime + 146097L }
  while (nyr >= 100L) { nyr <- nyr - 100L; dtime <- dtime + 36524L  }
  while (nyr >= 4L)   { nyr <- nyr - 4L;   dtime <- dtime + 1461L   }

  dtime <- dtime + (365L * nyr) + nmon[nmo + 1L]  # +1: R is 1-indexed
  if (nyr == 3L && nmo > 2L) dtime <- dtime + 1L
  dtime
}

# old_solarjd(): day + UT string (HHMM) → fractional MJD
old_solarjd <- function(day, mo, yr, ut) {
  dtime0 <- julian_zero(mo, yr)
  nhour  <- as.integer(substr(str_pad(ut, 4, pad = "0"), 1, 2))
  nmin   <- as.integer(substr(str_pad(ut, 4, pad = "0"), 3, 2))
  dday   <- ((60.0 * nhour) + nmin) / 1440.0
  dday + day + dtime0 - 2400000.5
}

# Vectorised wrapper for use inside mutate()
solar_jd_vec <- function(day, ut, mo = month, yr = year) {
  mapply(old_solarjd, day, mo, yr, ut)
}
```

---

## Compute derived quantities {#sec-compute}

```{r}
#| label: compute

results <- obs |>
  left_join(obsconst, by = "Obs") |>
  mutate(
    # Fill any observers absent from obsconst
    name  = coalesce(name,  "OBS NOT IN OBSCONST!"),
    k     = coalesce(k,     K_DEFAULT),
    Scale = coalesce(Scale, W_DEFAULT),

    # Julian date
    JD = solar_jd_vec(day, UT),

    # Wolf number sanity check
    count_calc = s + 10L * g,
    W_match    = (count_calc == W)
  )
```

### Wolf number mismatch check

```{r}
#| label: mismatch-check

mismatches <- filter(results, !W_match)
n_mm <- nrow(mismatches)

if (n_mm > 0L) {
  cat(sprintf("WARNING: %d observation(s) where submitted W ≠ s + 10g:\n", n_mm))
  mismatches |>
    select(uid, Obs, day, g, s, W, count_calc) |>
    kbl(booktabs = TRUE) |>
    kable_styling(latex_options = "hold_position", full_width = FALSE) |>
    print()
} else {
  cat("All W values match s + 10g. No mismatches found.\n")
}
```

### Calculate kc, log(c), log(kc)

```{r}
#| label: compute-logs

results <- results |>
  mutate(
    # k-corrected Wolf number (4 d.p., matching original)
    kcount = round(W * k, 4),

    # Log10; NA → sentinel string "-.----"
    logcount  = if_else(W      > 0, round(log10(W),      4), NA_real_),
    logkcount = if_else(kcount > 0, round(log10(kcount), 4), NA_real_),

    lc_str = if_else(
      is.na(logcount),
      "-.----",
      str_pad(formatC(logcount,  format = "f", digits = 4), 6, side = "right", pad = "0")
    ),
    lk_str = if_else(
      is.na(logkcount),
      "-.----",
      str_pad(formatC(logkcount, format = "f", digits = 4), 6, side = "right", pad = "0")
    ),

    # Name truncated to 20 characters (matches original printf field width)
    name_short = str_trunc(name, 20, ellipsis = "")
  ) |>
  arrange(JD)
```

### Preview

```{r}
#| label: tbl-results-preview
#| tbl-cap: "First 10 rows of computed results (sorted by JD)"

results |>
  slice_head(n = 10) |>
  select(JD, Obs, day, g, s, W, k, kcount, lc_str, lk_str, Scale, see) |>
  kbl(booktabs = TRUE, digits = 4, linesep = "") |>
  kable_styling(latex_options = c("striped","hold_position","scale_down"),
                full_width    = FALSE)
```

---

## Write `soldata.srt` {#sec-write}

Output format matches the original Perl `printf` exactly so downstream
tools that consume `soldata.srt` are unaffected.

```{r}
#| label: write-srt

output_file <- file.path(WD, "soldata.srt")

header_line <- "   JD         g    s   c     kc   log(c) log(kc)  w   Obs.      Obs. Name"

lines <- results |>
  mutate(
    line = sprintf(
      "%10.4f %4i %4i %4i %7.2f %6s %6s %5.3f %-4s %-20s   | %s",
      JD, g, s, W, kcount, lc_str, lk_str, Scale, Obs, name_short, see
    )
  ) |>
  pull(line)

writeLines(c(header_line, lines), con = output_file)
cat(sprintf("Wrote %d data line(s) to: %s\n", length(lines), output_file))
```

---

## Mark observations as published {#sec-publish}

A single `UPDATE … WHERE uid IN (…)` replaces the original row-by-row loop.

```{r}
#| label: mark-published

publish_uids <- obs$uid
uid_list     <- paste(publish_uids, collapse = ",")
sql_publish  <- sprintf(
  "UPDATE sun_data SET published = 1 WHERE uid IN (%s)", uid_list
)
rows_updated <- dbExecute(dbh, sql_publish)
cat(sprintf("Marked %d observation(s) as published.\n", rows_updated))
```

---

## Disconnect {#sec-disconnect}

```{r}
#| label: db-disconnect

dbDisconnect(dbh)
cat("Database connection closed.\n")
```

---

## Session info

```{r}
#| label: session-info
#| code-fold: true

sessionInfo()
```
