---
title: "AAVSO Solar Raw Processing"
subtitle: "Tidyverse port of RAWPROC_DB.PL v1.0 — Matthew Templeton, AAVSO HQ (2012-06-13)"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    self-contained: true
  pdf:
    toc: true
execute:
  echo: false
  warning: false
  message: false
params:
  month: 5        # Integer 1–12
  year:  2026     # Four-digit year ≥ 1940
  output_dir: "."
---

```{r setup, include=FALSE}
library(DBI)
library(RMariaDB)
library(tidyverse)
library(knitr)
library(kableExtra)
library(glue)
```

```{r constants}
K_DEFAULT <- 1.0   # fallback k-value when observer has none
W_DEFAULT <- 0.1   # fallback Scale/weight

months_lut  <- c("XXX","Jan","Feb","Mar","Apr","May","Jun",
                 "Jul","Aug","Sep","Oct","Nov","Dec")

stopifnot(params$month >= 1L, params$month <= 12L, params$year >= 1940L)

monthyr_str <- paste0(months_lut[params$month], " ", params$year)
```

```{r connect}
dbh <- dbConnect(
  RMariaDB::MariaDB(),
  dbname   = "aavso",
  host     = "localhost",
  user     = "root",
  password = "dJj135790",
  port     = 3333
)
```

## Helper functions

Ported directly from the Perl `old_solarjd` / `julian_zero` subroutines so
that Julian Date values in `soldata.srt` are byte-for-byte identical to those
produced by the original script.

```{r jd-helpers}
# Cumulative days before the start of each month (non-leap year, month 1..12).
# Matches Perl's @nmon array accessed as $nmon[$month].
.nmon <- c(0L, 31L, 59L, 90L, 120L, 151L, 181L, 212L, 243L, 273L, 304L, 334L)

# Port of Perl julian_zero($month, $year)
julian_zero <- function(month, year) {
  nyr   <- as.integer(year)
  if (nyr < 100L) nyr <- nyr + 2000L
  nyr   <- nyr - 1601L
  dtime <- 2305813L
  while (nyr >= 400L) { nyr <- nyr - 400L; dtime <- dtime + 146097L }
  while (nyr >= 100L) { nyr <- nyr - 100L; dtime <- dtime + 36524L  }
  while (nyr >= 4L)   { nyr <- nyr - 4L;   dtime <- dtime + 1461L   }
  dtime <- dtime + 365L * nyr + .nmon[month]
  if (nyr == 3L && month > 2L) dtime <- dtime + 1L
  dtime
}

# Port of Perl old_solarjd($day, $month, $year, $ut)
# ut is a 4-character string "HHMM"; numeric values are zero-padded automatically.
# Returns Modified Julian Date (MJD = JD - 2,400,000.5).
old_solarjd <- function(day, month, year, ut) {
  ut_str <- str_pad(as.character(ut), 4L, side = "left", pad = "0")
  nhour  <- as.integer(substr(ut_str, 1L, 2L))
  nmin   <- as.integer(substr(ut_str, 3L, 4L))
  frac   <- ((60L * nhour) + nmin) / 1440.0
  frac + day + julian_zero(month, year) - 2400000.5
}

# Round-half-up to `digits` decimal places.
# Matches Perl: int(10000 * x + 0.5) / 10000
round_half_up <- function(x, digits = 0L) {
  floor(x * 10^digits + 0.5) / 10^digits
}

# Format a log value as a 6-character string matching Perl's output.
# Returns "-.----" when the input is NA (i.e., count == 0).
fmt_log <- function(x) {
  if_else(
    is.na(x),
    "-.----",
    formatC(x, format = "f", digits = 4L, width = 6L)
  )
}
```

## Fetch observations

```{r fetch-obs}
sun_data <- tbl(dbh, "sun_data") |>
  filter(
    monthyr == monthyr_str,
    revised == 0L,
    !Obs %in% c("BSJ1", "TMT", "WEO", "SSTW", "ARUB")
  ) |>
  collect() |>
  arrange(Obs, day, UT)

publish_uids <- sun_data$uid   # save before any joins

n_data      <- nrow(sun_data)
n_observers <- n_distinct(sun_data$Obs)

cat(sprintf("There were %d observations by %d observers.\n", n_data, n_observers))

if (n_data == 0L) {
  cat("No data found for", monthyr_str, "— nothing to do.\n")
  dbDisconnect(dbh)
  knitr::knit_exit()
}
```

## Fetch observer constants

```{r fetch-obsconst}
olist <- unique(sun_data$Obs)

obsconst <- tbl(dbh, "sun_obsconst") |>
  filter(Obs %in% olist) |>
  collect() |>
  select(Obs, k, Scale, name, updated) |>
  arrange(Obs, desc(updated)) |>
  slice_max(updated, by = Obs, n = 1L, with_ties = FALSE) |>
  mutate(
    k     = if_else(is.na(k)     | k     == 0, K_DEFAULT, k),
    Scale = if_else(is.na(Scale) | Scale == 0, W_DEFAULT, Scale),
    name  = str_sub(coalesce(name, "OBS NOT IN OBSCONST!"), 1L, 20L)
  )

kbl(
  obsconst |> select(Obs, name, k, Scale),
  caption  = glue("Observer constants used for {monthyr_str}"),
  digits   = 4,
  booktabs = TRUE
) |>
  kable_styling(full_width = FALSE)
```

## Compute derived quantities {#sec-compute}

```{r compute}
results <- sun_data |>
  left_join(obsconst |> select(Obs, k, Scale, name), by = "Obs") |>
  mutate(
    k     = coalesce(k,     K_DEFAULT),
    Scale = coalesce(Scale, W_DEFAULT),
    name  = str_sub(coalesce(name, "OBS NOT IN OBSCONST!"), 1L, 20L)
  ) |>
  rowwise() |>
  mutate(JD = old_solarjd(day, params$month, params$year, UT)) |>
  ungroup() |>
  mutate(
    count      = s + 10L * g,
    W_match    = (count == W),

    # k-corrected Wolf number — round-half-up to match Perl
    kcount     = round_half_up(count * k, 4L),

    # Log10 values; NA sentinel when count == 0
    logcount   = if_else(count  > 0, round_half_up(log10(count),  4L), NA_real_),
    logkcount  = if_else(kcount > 0, round_half_up(log10(kcount), 4L), NA_real_),

    # 6-character string fields used in the printf output
    lc_str     = fmt_log(logcount),
    lk_str     = fmt_log(logkcount)
  ) |>
  arrange(JD)
```

### Wolf number mismatch check

```{r mismatch}
mismatches <- filter(results, !W_match)
n_mm <- nrow(mismatches)

if (n_mm > 0L) {
  cat(sprintf("WARNING: %d observation(s) where submitted W != s + 10*g:\n", n_mm))
  kbl(
    mismatches |> select(uid, Obs, day, g, s, W, count),
    booktabs = TRUE
  ) |>
    kable_styling(full_width = FALSE) |>
    print()
} else {
  cat("All W values match s + 10*g. No mismatches found.\n")
}
```

### Preview

```{r preview}
#| tbl-cap: "First 10 rows of computed results (sorted by MJD)"

results |>
  slice_head(n = 10L) |>
  select(JD, Obs, name, day, g, s, W, k, kcount, lc_str, lk_str, Scale, see) |>
  kbl(booktabs = TRUE, digits = 4, linesep = "") |>
  kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    full_width    = FALSE
  )
```

## Write `soldata.srt` {#sec-write}

Output format matches the original Perl `printf` exactly so downstream
tools that consume `soldata.srt` are unaffected.

```{r write-srt}
output_file <- file.path(params$output_dir, "soldata.srt")

header_line <- "   JD         g    s   c     kc   log(c) log(kc)  w   Obs.      Obs. Name"

lines <- results |>
  mutate(
    line = sprintf(
      "%10.4f %4i %4i %4i %7.2f %6s %6s %5.3f %-4s %-20s   | %s",
      JD,
      as.integer(g),
      as.integer(s),
      as.integer(W),
      kcount,
      lc_str,
      lk_str,
      Scale,
      Obs,
      name,
      see
    )
  ) |>
  pull(line)

writeLines(c(header_line, lines), output_file)
cat(sprintf("Wrote %d data line(s) to: %s\n", length(lines), output_file))
```

## Mark observations as published

```{r mark-published}
cat("Updating sun_data to mark observations as published.... please wait.\n")

walk(
  publish_uids,
  ~ dbExecute(dbh, "UPDATE sun_data SET published = 1 WHERE uid = ?", params = list(.x))
)

cat(sprintf("Marked %d observation(s) as published.\n", length(publish_uids)))
```

```{r disconnect}
dbDisconnect(dbh)
cat("Database connection closed.\n")
```
