---
title: "AAVSO Solar Raw Processing"
subtitle: "Port of RAWPROC_DB.PL v1.0 — Matthew Templeton, AAVSO HQ (2012-06-13)"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    self-contained: true
  pdf:
    toc: true
execute:
  echo: false
  warning: false
  message: false
params:
  month: 4        # Integer 1–12
  year:  2026     # Four-digit year ≥ 1940
  output_dir: "."  # Directory for soldata.srt; set to your WD if needed
---
```{r}
# Load Libraries
#
#| label: setup
#| include: false
rm(list = ls())
library(DBI)
library(RMariaDB)      
library(tidyverse)
library(knitr)
library(kableExtra)
library(lubridate)
library(glue)
library(writexl)
```

```{r}
#| label: db-connect

dbh <- dbConnect(
  RMariaDB::MariaDB(),
  dbname   = "aavso",
  host     = "localhost",
  user     = "root",
  password = "dJj135790",
  port     = 3333
)
```

# Grab Data

```{r}
#| label: defaults

K_DEFAULT <- 1.0
W_DEFAULT <- 0.1   # "Scale" / weight
```

```{r}
# Pull April 2206 Observations
sun_data <- tbl(dbh,"sun_data") |> collect() |> filter(monthyr =="Apr 2026")
```
```{r}
# Create table with one row per Obs
sun_header <- tbl(dbh,"sun_header") |> collect()
observers <- sun_header |>
  select(Obs, name, updated,method) |>
  arrange(Obs, desc(updated)) |>
  slice_max(updated, by = Obs, n = 1, with_ties = FALSE)


n_obs       <- nrow(observers)
n_observers <- n_distinct(observers$Obs)
cat(sprintf("Retrieved %d observations from %d observer(s).\n",
            n_obs, n_observers))

# Early exit if no data — stops rendering cleanly without an error page
if (n_obs == 0L) {
  cat("No data found for", monthyr, "— nothing to do.\n")
  dbDisconnect(dbh)
  knitr::knit_exit()
}

```

```{r}
# Create Table of Lat K Value
obsconst <- tbl(dbh,"sun_obsconst") |> collect()
kvalues <- obsconst |>
  select(Obs,k,updated,Scale) |>
  arrange(Obs, desc(updated)) |>
  slice_max(updated, by = Obs, n = 1, with_ties = FALSE)

# Apply defaults for missing / zero values
kvalues <- kvalues |>
  mutate(
    k     = if_else(is.na(k)     | k     == 0, K_DEFAULT, k),
    Scale = if_else(is.na(Scale) | Scale == 0, W_DEFAULT, Scale)
  )

# dbWriteTable(dbh,"kvalues",kvalues, overwrite=TRUE)
```

## Julian Date helpers {#sec-jd}

Direct port of the original JULIAN0 QuickBasic algorithm used in the Perl
script. Results are Modified Julian Dates (MJD).
```{r}
#| label: jd-functions

# # julian_zero(): MJD base for day 0 of the given month/year
# julian_zero <- function(nmo, nyear) {
#   nmon <- c(0L, 0L, 31L, 59L, 90L, 120L, 151L, 181L, 212L, 243L, 273L, 304L, 334L)
#   nyr  <- if (nyear < 100L) nyear + 2000L else nyear
#   nyr  <- nyr - 1601L
# 
#   dtime <- 2305813L
#   while (nyr >= 400L) { nyr <- nyr - 400L; dtime <- dtime + 146097L }
#   while (nyr >= 100L) { nyr <- nyr - 100L; dtime <- dtime + 36524L  }
#   while (nyr >= 4L)   { nyr <- nyr - 4L;   dtime <- dtime + 1461L   }
# 
#   dtime <- dtime + (365L * nyr) + nmon[nmo + 1L]  # +1: R is 1-indexed
#   if (nyr == 3L && nmo > 2L) dtime <- dtime + 1L
#   dtime
# }
# 
# # old_solarjd(): day + UT string (HHMM) -> fractional MJD
# old_solarjd <- function(day, mo, yr, ut) {
#   dtime0 <- julian_zero(mo, yr)
#   ut_pad <- str_pad(ut, 4, pad = "0")
#   nhour  <- as.integer(substr(ut_pad, 1, 2))
#   nmin   <- as.integer(substr(ut_pad, 3, 4))   # fixed: was (3, 2)
#   dday   <- ((60.0 * nhour) + nmin) / 1440.0
#   dday + day + dtime0 - 2400000.5
# }
# 
# # Vectorised wrapper for use inside mutate()
# solar_jd_vec <- function(day, ut, mo = month, yr = year) {
#   mapply(old_solarjd, day, mo, yr, ut)
# }
```

## Compute derived quantities {#sec-compute}

```{r}
#| label: compute
sun_data$date <- as.numeric(sun_data$date)
results <- sun_data |>
  inner_join(kvalues,by="Obs") |>
  mutate(
    # Fill any observers absent from obsconst
  #  name  = coalesce(name,  "OBS NOT IN OBSCONST!"),
    k     = coalesce(k,     K_DEFAULT),
    Scale = coalesce(Scale, W_DEFAULT),

    # Julian date
    JD = as_date(date - 2440587.5),
 #   JD = solar_jd_vec(day, UT),

    # Wolf number sanity check
    count_calc = s + 10L * g,
    W_match    = (count_calc == W) 
  )
```

```{r}
# Monthlyr Raw Sun


 monthly_raw <- sun_data |> group_by(day) |> summarize(
      min_w = min(W,na.rm=TRUE),
      max_w = max(W,na.rm=TRUE),
      mean_w = mean(W,na.rm=TRUE),
      med_w = median(W,na.rm=TRUE),
      row_cnt = n(),
      # Spots
      min_s = min(s,na.rm=TRUE),
      max_s = max(s,na.rm=TRUE),
      mean_s = mean(s,na.rm=TRUE),
      med_s = median(s,na.rm=TRUE),
      # Groups
      min_g = min(g,na.rm=TRUE),
      max_g = max(g,na.rm=TRUE),
      mean_g = mean(g,na.rm=TRUE),
      med_g = median(g,na.rm=TRUE)
      
  ) |> ungroup()
```
## Calculate Monthly Numbers using Monthlyr column
```{r}
monthly_raw <- sun_data |> group_by(monthyr) |> summarize(
    min_w = min(W, na.rm = TRUE),
    max_w = max(W, na.rm = TRUE),
    mean_w = mean(W, na.rm = TRUE),
    med_w = median(W, na.rm = TRUE),
    row_cnt = n(),
    # Spots
    min_s = min(s, na.rm = TRUE),
    max_s = max(s, na.rm = TRUE),
    mean_s = mean(s, na.rm = TRUE),
    med_s = median(s, na.rm = TRUE),
    # Groups
    min_g = min(g, na.rm = TRUE),
    max_g = max(g, na.rm = TRUE),
    mean_g = mean(g, na.rm = TRUE),
    med_g = median(g, na.rm = TRUE)

) |> ungroup()
```

## Calcuate using standard date column

```{r}
monthly_raw <- sun_data |>
    mutate(month = floor_date(JD, "month")) |>
    group_by(month) |>
    summarize(
        # Raw Wolf Numbers
        min_w = min(W, na.rm = TRUE),
        max_w = max(W, na.rm = TRUE),
        mean_w = mean(W, na.rm = TRUE),
        med_w = median(W, na.rm = TRUE),
        row_cnt = n(),
        # Spots
        min_s = min(s, na.rm = TRUE),
        max_s = max(s, na.rm = TRUE),
        mean_s = mean(s, na.rm = TRUE),
        med_s = median(s, na.rm = TRUE),
        # Groups
        min_g = min(g, na.rm = TRUE),
        max_g = max(g, na.rm = TRUE),
        mean_g = mean(g, na.rm = TRUE),
        med_g = median(g, na.rm = TRUE)

    ) |> ungroup()
``


### Wolf number mismatch check

```{r}
#| label: mismatch-check

mismatches <- filter(results, !W_match)
n_mm <- nrow(mismatches)

if (n_mm > 0L) {
  cat(sprintf("WARNING: %d observation(s) where submitted W != s + 10g:\n", n_mm))
  mismatches |>
    select(uid, Obs, day, g, s, W, count_calc) |>
    kbl(booktabs = TRUE) |>
    kable_styling(latex_options = "hold_position", full_width = FALSE) |>
    print()
} else {
  cat("All W values match s + 10g. No mismatches found.\n")
}
```

### Calculate kc, log(c), log(kc)

```{r}
#| label: compute-logs

results <- results |>
  mutate(
    # k-corrected Wolf number (4 d.p., matching original)
    kcount = round(W * k, 4),

    # Log10; NA -> sentinel string "-.----"
    logcount  = if_else(W      > 0, round(log10(W),      4), NA_real_),
    logkcount = if_else(kcount > 0, round(log10(kcount), 4), NA_real_),

    # formatC width=6 right-justifies to match %6s in the output sprintf
    lc_str = if_else(
      is.na(logcount),
      "-.----",
      formatC(logcount,  format = "f", digits = 4, width = 6)
    ),
    lk_str = if_else(
      is.na(logkcount),
      "-.----",
      formatC(logkcount, format = "f", digits = 4, width = 6)
    ),

    # Name truncated to 20 characters (matches original printf field width)
   # name_short = str_trunc(name, 20, ellipsis = "") 
  ) |>
  arrange(JD)

results <- results |>
  inner_join(observers) |>
  mutate(name_short = str_trunc(name, 20, ellipsis = ""))

```

### Preview

```{r}
#| label: tbl-results-preview
#| tbl-cap: "First 10 rows of computed results (sorted by JD)"

results |>
  slice_head(n = 10) |>
  select(JD,name_short, Obs, day, g, s, W, k, kcount, lc_str, lk_str, Scale, see) |>
  kbl(booktabs = TRUE, digits = 4, linesep = "") |>
  kable_styling(latex_options = c("striped","hold_position","scale_down"),
                full_width    = FALSE)
```
## Write `soldata.srt` {#sec-write}

Output format matches the original Perl `printf` exactly so downstream
tools that consume `soldata.srt` are unaffected.

```{r}
#| label: write-srt



output_file <- file.path(getwd(), "soldata.srt")
header_line <- "   JD         g    s   c     kc   log(c) log(kc)  w   Obs.      Obs. Name"

results |>
  mutate(
    line = glue(
      "{formatC(JD, format='f', digits=4, width=10)} ",
      "{formatC(g,  format='d', width=4)} ",
      "{formatC(s,  format='d', width=4)} ",
      "{formatC(W,  format='d', width=4)} ",
      "{formatC(kcount, format='f', digits=2, width=7)} ",
      "{str_pad(lc_str,  6, 'left')} ",
      "{str_pad(lk_str,  6, 'left')} ",
      "{formatC(Scale, format='f', digits=3, width=5)} ",
      "{str_pad(Obs,   4, 'right')} ",
      "{str_pad(name_short, 20, 'right')}   | {see}",
      .sep = ""
    )
  ) |>
  pull(line) |>
  (\(lines) {
    writeLines(c(header_line, lines), con = output_file)
    cat(sprintf("Wrote %d data line(s) to: %s\n", length(lines), output_file))
  })()

# Write Results Data.frame to Excel sheet

results |> select(JD,g,s,W,kcount,lc_str,lk_str,Scale,Obs,name_short,see) |> write_xlsx("results.xlsx")
```

## Plot Monthly Raw Numbers
### Wolf

```{r}
ggplot(monthly_raw) + 
  geom_line(aes(x=day, y=min_w, colour="Min"), linewidth=2) +
  geom_line(aes(x=day, y=max_w, colour="Max"), linewidth=2) +
  geom_line(aes(x=day, y=mean_w, colour="Mean"), linewidth=2) + 
  theme_light() +
  labs(title="Minimum, Average, and Maximum Raw W (wolf)", y="Raw Wolf") +
  theme(legend.position = "bottom")
``` 

### Spots

```{r}
ggplot(monthly_raw) + 
  geom_area(aes(x=day, y=min_s, colour="Min",fill="Min"), linewidth=2) +
  geom_area(aes(x=day, y=max_s, colour="Max"), linewidth=2) +
  geom_area(aes(x=day, y=mean_s, colour="Mean"), linewidth=2) + 
  theme_light() +
  labs(title="Minimum, Average, and Maximum Spots", y="Spots") +
  theme(legend.position = "bottom")
``` 
### Spots

```{r}
ggplot(monthly_raw) + 
  geom_line(aes(x=day, y=min_g, colour="Min"), linewidth=2) +
  geom_line(aes(x=day, y=max_g, colour="Max"), linewidth=2) +
  geom_line(aes(x=day, y=mean_g, colour="Mean"), linewidth=2) + 
  theme_light() +
  labs(title="Minimum, Average, and Maximum Groups", y="Groups") +
  theme(legend.position = "bottom")
``` 